---
title: "Network data in  R"
output: html_notebook
---

# Introduction
This notebook accompanies the second practical P2 of the Barcelona Past Networks Summer School. It shows how to create basic network in R.

# Preliminaries 
For this network we will need the package igraph.
```{r list required packages}
required_packages <- c("igraph")
```

We install the packages that we have not installed yet.
```{r install required packages}
packages_to_install <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(packages_to_install)) install.packages(packages_to_install)
```

Finally, we load the packages
```{r install packages}
invisible(lapply(required_packages, library, character.only = TRUE))
```

Here, we use the following versions of the packages:
* igraph:   v1.4.3

We also use R version 4.3.0

# Exercise on creating a basic network 
First of all we record the info as an edge list
```{r edgelist}
# record info in an edgelist
el <- matrix(c(c("v1","v1","v2"), # these are the sources 
               c("v2","v3","v1")),# these are the targets
             ncol=2)

# convert into a graph
el_g <- graph_from_edgelist(el, directed = TRUE)
V(el_g) # view the nodes
E(el_g) # view the edges

# plot the graph
plot(el_g)
```

Convert this into an adjacency list
```{r adjacency edge list}
al <- as_adj_edge_list(el_g)
```

Write the network as an adjacency matrix
```{r adjacency matrix}
# define the adjacency matrix of the network
am <- matrix(c(0,1,1,1,0,0,0,0,0), ncol = 3, byrow = TRUE)
rownames(am) <- c("v1","v2","v3")
colnames(am) <- c("v1","v2","v3")

# create a network from the adjacency matrix
am_g <- graph_from_adjacency_matrix(am)
```

Check that the two networks are identical
```{r identical graphs}
# check that the graphs are identical
identical_graphs(am_g,el_g)
```

Plot the two networks one next to the other for a visual confirmation
```{r plot identical graphs together}
# plot them together
par(mfrow=c(1,2))
plot(el_g, main = "Network from the edge list")
plot(am_g, main = "Network from the adjacency matrix")
```

Add a node v4
```{r add a node v4}
# add vertex to the graph
el_2_g<- add.vertices(el_g,1,name="v4")
# define edgelist
el_2 <- as_edgelist(el_2_g)
# compare with original edgelist
all(el==el_2)

# add vertex to adjacency matrix
am_2 <- rbind(am,v4=0)
am_2 <- cbind(am_2,v4=0)
# create network out of it
am_2_g <- graph_from_adjacency_matrix(am_2)

# compare the modified networks
identical_graphs(am_2_g,el_2_g)
```

# Exercise 1 on attributes
Assign size and weight attributes
```{r assign size and weight attributes}
# define attribute dataframes
node_attributes_df <- data.frame(name = c("v1","v2","v3"), size = c(10,20,30))
edge_attributes_df <- data.frame(weight = c(1,10,1))

# assign attributes
V(am_g)$size <- node_attributes_df$size
E(am_g)$weight <- edge_attributes_df$weight

# check that they have been assigned
vertex_attr(am_g)
edge_attr(am_g)

# plot the graph
plot(am_g, 
     vertex.size = V(am_g)$size,
     edge.width = E(am_g)$weight)
```

# Exercise 2 on attributes
Retrieve dataframe city_pop_df
```{r define the dataframe city_pop_df}
# define columns of the dataframe 
city <- c("Barcelona","L'Hospitalet de Llobregat","Terrassa","Badalona")
pop <- c(1664182,269382,223627,223166)
lon <- c(2.176944, 2.1, 2.013, 2.246111)
lat <- c(41.382778, 41.359722, 41.57, 41.448889)

# define dataframe
city_pop_df <- data.frame(city,pop,lon,lat)

# initialise "category" column 
city_pop_df$category <- NA

# define function to associate a category to the population
which_category <- function(x){
  try(if(!is.numeric(x)) stop("The input must be a numerical value"))
  if (x >= 1000000){category <- "large"}
  else if (x < 1000000 & x>= 250000){category <- "medium"}
  else if (x < 250000){category <- "small"}
  return(category)
}

# assign a category to each city using the above function
for (i in 1:nrow(city_pop_df)) {
  city_pop_df$category[i] <- which_category(city_pop_df$pop[i])
}

# transform the column into factors
city_pop_df$category <- factor(city_pop_df$category)
```

Create network of cities based on distances
```{r create network of cities within a certain distance}
# find number of cities
n_cities <- nrow(city_pop_df)

# initialise matrix of longitude differences
lon_diff <- matrix(NA, 
                   ncol = n_cities,
                   nrow = n_cities)
colnames(lon_diff) <- city_pop_df$city
rownames(lon_diff) <- city_pop_df$city

# initialise matrix of latitude differences
lat_diff <- lon_diff

# check that pairs of nodes
for (i in 1:n_cities) {
  for (j in 1:n_cities) {
    lon_diff[i,j] <- abs(city_pop_df[i,"lon"]-city_pop_df[j,"lon"])
    lat_diff[i,j] <- abs(city_pop_df[i,"lat"]-city_pop_df[j,"lat"])
    lon_diff[i,i] <- Inf # ensure that the cities are not connected to themselves
  }
}

# derive adj matrix
adj_matrix <- (lat_diff<0.1) * (lon_diff<0.1)

# define network out of it
city_network <- graph_from_adjacency_matrix(adj_matrix, mode = "undirected")

# assign attributes
V(city_network)$city <- city_pop_df$city
V(city_network)$pop <- city_pop_df$pop
V(city_network)$lon <- city_pop_df$lon
V(city_network)$lat <- city_pop_df$lat
V(city_network)$category <- as.character(city_pop_df$category)

# plot the network
plot(city_network, 
     vertex.size = sqrt(V(city_network)$pop/5000),
     vertex.color = as.numeric(V(city_network)$category)+1,
     layout = cbind(V(city_network)$lon,V(city_network)$lat)
     )
```

Export the data
```{r export the data}
# save as graphml
write_graph(city_network,
            "data/city_network.graphml", #this saves the network in your working directory.
            format="graphml")
```
