---
title: "P3 network analysis"
output: html_notebook
---

# Introduction 
Through this notebook, we will learn the basics of network analysis in R. 
It complements the Practical P3 in the Barcelona Past Networks Summer School.
The notebook is largely based on Chapter 4 of

Peeples, Matthew A. and Tom Brughmans (2023). 
Online Companion to Network Science in Archaeology. 
https://archnetworks.net, Accessed 2024-03-07. 

# Preliminaries
First of all, we list the packages we will need:
```{r list required packages}
required_packages <- c("igraph", 
                       "statnet", 
                       "intergraph", 
                       "vegan", 
                       "sna")
```

We install the packages that we have not installed yet
```{r install required packages}
packages_to_install <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(packages_to_install)) install.packages(packages_to_install)
```

Finally, we load the packages
```{r install packages}
invisible(lapply(required_packages, library, character.only = TRUE))
```

Here, we use the following versions of the packages:
* igraph:     v1.4.3
* statnet:    v2019.6
* intergraph: v2.0.3
* vegan:      v2.6.4
* sna:        v2.7.1

This code was written in R 4.3.0

# Basics 
```{r Handle basic networks}
# define edge list matrix
example_el <- matrix(c("v1","v1","v2", # these are the sources 
                       "v2","v3","v3"),# these are the targets 
                     ncol=2)

# build network from edgelist
example_g <- graph_from_edgelist(example_el, directed = TRUE)

# plot
plot(example_g)

# access vertices/nodes
V(example_g)

# access edges
E(example_g)

# add labels to edges
E(example_g)$label <- c("A","B","C")

# add size attributes to node
V(example_g)$size <- c(10,20,30)

# view vertex attributes now
vertex.attributes(example_g)

# the basic plot function automatically interprets the size and label attributes
plot(example_g)
```

# Data
Load the necessary data. Note that the paths point to a data folder contained
within the working directory of the project. If the data is located elsewhere,
you will need to specify the correct path.

```{r data}
# simple net
cibola_adj_mat <- read.csv(file = "data/Cibola_adj.csv", # path to the data 
                           header = T,
                           row.names = 1)

# Simple network with isolates
simple_net <-
  graph_from_adjacency_matrix(
    as.matrix(cibola_adj_mat),
    mode = "undirected")

# weighted net
cibola_edgelist <-
  read.csv(file = "data/Cibola_edgelist.csv", header = TRUE)

# Create a weighted undirected network by adding column of random
# weights to edge list
cibola_edgelist$weight <- sample(
  seq(1, 4),
  nrow(cibola_edgelist),
  replace = TRUE)

# initialise network
weighted_net <-
  igraph::graph_from_edgelist(as.matrix(cibola_edgelist[, 1:2]),
                              directed = FALSE)

# assign weights to network
E(weighted_net)$weight <- cibola_edgelist$weight
```

Check what they look like in simple plots
```{r plot of loaded networks}
# plot "original" network
plot(simple_net, main = "Cibola network")

# plot them together
par(mfrow=c(1,2))
set.seed(3523) # Set random seed to ensure graph layout stays the same each time.
plot(simple_net, main = "Cibola network, simple net")
set.seed(3523)
plot(weighted_net, edge.width = E(weighted_net)$weight, main ="Weighted Cibola network")
```
Remark that simple_net contains an isolated node that does not appear in weighted_net.
This is because weighted_net is created from an edgelist, a network data format
that does not retain information on isolated nodes.

# Example on analysis
First of all, find basic global measures
```{r basic lobal measures}
# number of nodes
vcount(simple_net)
vcount(weighted_net)

# number of edges
ecount(simple_net)
ecount(weighted_net)

# density
graph.density(simple_net)
graph.density(weighted_net)
```

# Exercise on paths and connectivity
```{r paths and distances}
# find neighbours of a node
neighbors(simple_net, "Atsinna")

# distances between specific nodes
distances(simple_net,v="Atsinna",to="UG494")   # unweighted
distances(weighted_net,v="Atsinna",to="UG494") # weighted

# find a shortest path
shortest_paths(simple_net,from="Atsinna",to="UG494") #unweighted (steps)
shortest_paths(weighted_net,from="Atsinna",to="UG494") # weighted

# find all shortest paths
all_shortest_paths(simple_net,from="Atsinna",to="UG494") #unweighted (steps)
all_shortest_paths(weighted_net,from="Atsinna",to="UG494") # weighted

# find weight of edges along a shortest path from Atsinna to UG494 in the weighted network
Atsinna_UG494_epath <- shortest_paths(weighted_net,
                                      from="Atsinna",
                                      to="UG494",
                                      output ="epath")$epath[[1]]
E(weighted_net)[Atsinna_UG494_epath]$weight

# longest shortest path
farthest_vertices(simple_net)
farthest_vertices(weighted_net)

# length of longest shortest path
diameter(simple_net)
diameter(weighted_net)

# connectivity
is_connected(simple_net) #simple_net is based on the contains 
is_connected(weighted_net)

# components
decompose(simple_net)
decompose(weighted_net)

# number of components
length(decompose(simple_net))
components(cibola_adj_mat) # dedicated function in sna, requires the adjacency matrix

# bridges
bridges(simple_net)
bridges(weighted_net)

# cutpoints
cutpoints(cibola_adj_mat) # dedicated function in sna, requires the adjacency matrix
```

# Exercise on centrality
```{r centrality}
# degree 
simple_net_degree <- igraph::degree(simple_net)
weighted_net_degree <- igraph::degree(weighted_net)

# strength 
simple_net_strength <- igraph::strength(simple_net)
weighted_net_strength <- igraph::strength(weighted_net)

# betweenness
simple_net_betweenness <- igraph::betweenness(simple_net)
weighted_net_betweenness <- igraph::betweenness(weighted_net)

# closeness
simple_net_closeness <- igraph::closeness(simple_net)
weighted_net_closeness <- igraph::closeness(weighted_net)

# statistics dataframe
centrality_scores <- data.frame(
  networks = c("simple","weighted"),
  min_degree = c(
    min(simple_net_degree),
    min(weighted_net_degree)
    ),
  max_degree = c(
    max(simple_net_degree),
    max(weighted_net_degree)
    ),
  mean_degree = c(
    mean(simple_net_degree),
    mean(weighted_net_degree)
    ),
  median_degree = c(
    median(simple_net_degree),
    median(weighted_net_degree)
    ),
  centralisation_degree = c(
    centralization.degree(simple_net)$centralization,
    centralization.degree(weighted_net)$centralization
  ),
  min_betweenness = c(
    min(simple_net_betweenness),
    min(weighted_net_betweenness)
    ),
  max_betweenness = c(
    max(simple_net_betweenness),
    max(weighted_net_betweenness)
    ),
  mean_betweenness = c(
    mean(simple_net_betweenness),
    mean(weighted_net_betweenness)
    ),
  median_betweenness = c(
    median(simple_net_betweenness),
    median(weighted_net_betweenness)
    ),
  centralisation_betweenness = c(
    centralization.betweenness(simple_net)$centralization,
    centralization.betweenness(weighted_net)$centralization
  )
)

# degree distribution
simple_net_deg_distribution <- degree_distribution(simple_net)
hist(simple_net_degree,
     breaks = seq(-0.5, length(simple_net_deg_distribution)+0.5, by=1),
     main="Degree distribution in simple-net",
     xlab="degree",
     ylab="occurrences")
```

# Exercise on communities
```{r communities}
# find communities
gn <- igraph::edge.betweenness.community(simple_net) 
wt <- igraph::cluster_walktrap(simple_net, steps = 4)
lv <- igraph::cluster_louvain(simple_net) 

#modularity
simple_net_modularity <- data.frame(
  method = c("Grivan-Newmann","Walktrap Algorithm", "Louvain Method"),
  modularity = c(
    modularity(
      simple_net, 
      membership = membership(gn)),
    modularity(
      simple_net, 
      membership = membership(wt)),
    modularity(
      simple_net, 
      membership = membership(lv))
  )
)

# plots
par(mfrow=c(1,3))
set.seed(3523) # Set random seed to ensure graph layout stays the same each time.
plot(gn, simple_net, main = "Grivan-Newmann")
set.seed(3523)
plot(wt, simple_net, main = "Walktrap")
set.seed(3523)
plot(lv, simple_net, main = "Louvain")

# crossings
igraph::crossing(lv, simple_net) 
```

