---
title: "R Notebook"
output: html_notebook
---

# Introduction 
Through this notebook, we will learn the basics of network visualisation in R.

# Preliminaries
First of all, we list the packages we will need 
```{r list required packages}
required_packages <- c("igraph", "statnet", "intergraph", "vegan", "tidyverse","ggraph")
```

We install the packages that we have not installed yet
```{r install required packages}
packages_to_install <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(packages_to_install)) install.packages(packages_to_install)
```

Finally, we load the packages
```{r install packages}
invisible(lapply(required_packages, library, character.only = TRUE))
```

Here, we use the following versions of the packages:
* igraph:     v1.4.3
* statnet:    v2019.6
* intergraph: v2.0.3
* vegan:      v2.6.4
* tidyverse   v2.0.0
* ggraph      v2.1.0

# Data 
Load the necessary data
```{r networks from Cibola data}
# first, read the adjacency matrix
cibola_adj_mat <- read.csv(file = "data/networks/Cibola_adj.csv",
                           header = T,
                           row.names = 1)

# then read the network attributes
cibola_attr <- read.csv(file = "data/networks/Cibola_attr.csv", header = TRUE)


# build igraph
cibola_i <-
  igraph::graph_from_adjacency_matrix(as.matrix(cibola_adj_mat),
                                      mode = "undirected")

# build statnet/network
cibola_n <- asNetwork(cibola_i)
```

# Basic plots with igraph and statnet
Compare the behaviour of igraph and statnet
```{r compare igraph and statnet}
# plot together basic igraph and statnet
par(mfrow=c(1,2))
plot(cibola_i) # more information, less readable
plot(cibola_n) # less information, more sensible defaults
```

Add information on 

