---
title: "R Notebook"
output: html_notebook
---

# Introduction 
Through this notebook, we will learn the basics of network visualisation in R.
It complements the Practical P4 in the Barcelona Past Networks Summer School.
The notebook is largely based on Chapter 6 of

Peeples, Matthew A. and Tom Brughmans (2023). 
Online Companion to Network Science in Archaeology. 
https://archnetworks.net, Accessed 2024-03-07. 

# Preliminaries
First of all, we list the packages we will need 
```{r list required packages}
required_packages <- c("igraph", 
                       "statnet", 
                       "intergraph", 
                       "tidyverse",
                       "ggraph",
                       "ggforce",     # convex hull
                       "concaveman",  # convex hull
                       "sf",
                       "ggmap",
                       "sfnetwork"
                       )
```

We install the packages that we have not installed yet
```{r install required packages}
packages_to_install <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(packages_to_install)) install.packages(packages_to_install)
```

Finally, we load the packages
```{r install packages}
invisible(lapply(required_packages, library, character.only = TRUE))
```

Here, we use the following versions of the packages:
* igraph:     v1.4.3
* statnet:    v2019.6
* intergraph: v2.0.3
* vegan:      v2.6.4
* tidyverse:  v2.0.0
* ggraph:     v2.1.0
* ggforce:    v0.4.1 
* concaveman: v1.1.0
* sf:         v1.0.16
* ggmap:      v4.0.0
* sfnetwork:  v0.6.3

This code was written in R 4.3.0

# Data 
Before starting, please download the Cibola data from the Online Companion website https://book.archnetworks.net/dataandworkspace#Cibola
Place the data in the folder "data" within your project.

Load the necessary data
```{r networks from Cibola data}
# first, read the adjacency matrix
cibola_adj_mat <- read.csv(file = "data/Cibola_adj.csv",
                           header = T,
                           row.names = 1)

# then read the network attributes
cibola_attr <- read.csv(file = "data/Cibola_attr.csv", header = TRUE)


# build igraph
cibola_i <-
  igraph::graph_from_adjacency_matrix(as.matrix(cibola_adj_mat),
                                      mode = "undirected")

# attach info to the graphs
for (a in colnames(cibola_attr)) {
  vertex_attr(cibola_i,a) <- cibola_attr[[a]]
}

# build statnet/network
cibola_n <- asNetwork(cibola_i)
```

# Basic plots with igraph and statnet
Compare the behaviour of igraph and statnet
```{r compare igraph and statnet}
# plot together basic igraph and statnet
par(mfrow=c(1,2))
plot(cibola_i, main ="Cibola, igraph default") # more information, less readable
plot(cibola_n, main ="Cibola, statnet default") # less information, more sensible defaults
```

Fairly similar visualisation tools.
```{r visuals with igraph and statnet}
# plot them together
par(mfrow=c(1,2))

# plot igraph
plot(cibola_i, 
     vertex.size = igraph::degree(cibola_i),
     vertex.color = as.factor(cibola_attr$Region),
     edge.color = "darkgray",
     vertex.label = NA,
     main = "Cibola, igraph customised")

# plot statnet
plot(
  cibola_n,
  vertex.cex = sna::degree(cibola_n) / 10,
  vertex.col = as.factor(cibola_attr$Region),
  edge.col = "darkgray",
  main = "Cibola, statnet customised")
```

# Plots with ggraph
## Node/Edges
```{r basic}
# basic plot
ggraph(cibola_i) +
  # add edges
  geom_edge_link() +
  # add nodes
  geom_node_point()
```  

Remark: easy to show only edges/nodes
```{r only nodes}
# basic plot
ggraph(cibola_i) +
  # add nodes
  geom_node_point() +
  # select theme
  theme_graph()
```

```{r only edges}
# basic plot
ggraph(cibola_i) +
  # add nodes
  geom_edge_link() +
  # select theme
  theme_graph()
```

Node and Edge aesthetics can be defined for the whole graph or individually (using aes).
They can either be attributes or come from external values (e.g., computed using a function)
```{r size and colour}
# basic plot
ggraph(cibola_i) +
  # add edges
  geom_edge_link(
    size = "2",
    color = "gray"
  ) +
  # add nodes
  geom_node_point(
    aes(size = igraph::degree(cibola_i),
        color = Region)
  ) +
  # select theme
  theme_graph()
```

The links can be represented as straight or curved lines
```{r shape}
# basic plot
ggraph(cibola_i) +
  # add edges
  geom_edge_arc(
    size = "2",
    color = "gray",
    linetype = "dashed"
  ) +
  # add nodes
  geom_node_point(
    aes(size = igraph::degree(cibola_i),
        color = Region),
    shape = "square"
  ) +
  # select theme
  theme_graph()
```

We can add labels
```{r labels}
# base graph
ggraph(cibola_i) +
  # add links
  geom_edge_link(
    color ="gray"
  ) +
  # add nodes
  geom_node_point(
    aes(size = igraph::degree(cibola_i),
        color = Region)
  ) +
  # add labels
  geom_node_text(
    aes(label = name),
    repel=T
  ) + 
  # select theme
  theme_graph()
```

We can customise the layout by choosing from existing ones (e.g., circle) or defining our own
```{r layout circle}
# Define circular layout
ggraph(cibola_i, 
       layout = "circle") +
  geom_edge_link(
    color ="gray"
  ) +
  geom_node_point(
    aes(size = igraph::degree(cibola_i),
        color = Region)
  ) +
  geom_node_text(
    aes(label = name)
  ) + 
  # select theme
  theme_graph()
```

Alternatively, we can define a layout manually
```{r layout manual}
# Create a data frame for the custom layout
coord_df <- data.frame(
  name = V(cibola_i)$name,
  x = V(cibola_i)$x, 
  y = V(cibola_i)$y)

# Define layout manually
ggraph(cibola_i, 
       layout = "manual", 
       x = coord_df$x,
       y = coord_df$y) +
  geom_edge_link(
    color ="gray"
  ) +
  geom_node_point(
    aes(size = igraph::degree(cibola_i),
        color = Region)
  ) +
  geom_node_text(
    aes(label = name)
  ) + 
  # select theme
  theme_graph()
```

It is also possible to define a layout interactively.
First open the interactive layout (it will open a new window. 
```{r define layout interactively}
# uses tkplot in igraph
tkid <- tkplot(cibola_i) #tkid is the id of the tkplot that will open
```

Once you are happy, grab the coordinates of the layout
```{r grab coordinates}
l <- tkplot.getcoords(tkid) # grab the coordinates from tkplot
```

Then close the window.

Finally, we are ready to use the layout in the plot
```{r plots with modified layout}
# plot using igraph
plot(cibola_i, layout=l)

# plot using ggraph
ggraph(cibola_i, 
       layout = "manual",
       x = l[,1],
       y = l[,2]) +
  geom_edge_link(
    color ="gray"
  ) +
  geom_node_point(
    aes(size = igraph::degree(cibola_i),
        color = Region)
  ) +
  geom_node_text(
    aes(label = name)
  ) + 
  # select theme
  theme_graph()
```

## Hulls and maps
We have seen how the plot function with igraph and statnet allows to visualise a convex hull around nodes to identify them as part of a cluster. The same can be done with ggplot
```{r hull}
# identify the clusters
grp <- as.factor(cluster_louvain(cibola_i)$membership)

# start graph
ggraph(cibola_i, layout = "fr") +
  # add links
  geom_edge_link(width = 0.2) +
  # add points
  geom_node_point(aes(fill = grp),
                  shape = 21,
                  size = 5,
                  alpha = 0.75) +
  # Create hull around points within group and label
  geom_mark_hull(
    aes(
      x,
      y,
      group = grp,
      fill = grp,
    ),
    concavity = 4, # the larger it is the more convex
    expand = ggplot2::unit(2, "mm"),
    alpha = 0.25,
  ) +
  # define color scheme
  scale_fill_brewer(palette = "Set2") +
  theme_graph()

?scale_fill_brewer
```

## Maps.
The choice of an appropriate map depends on the scale and what one wants to highlight
For some cases, a terrain is best. This can be done via stadiamap.
Annoyingly, then one needs to add nodes and links as points and segments on the map.
```{r terrain map with stadiamap}
# need to register stadiamaps beforehand (see tutorial)
# ggmap::register_stadiamaps("HERE YOUR REGISTRATION CODE") # insert your registration code in the function

# download stadiamap
base_cibola <- get_stadiamap(
  bbox = c(-110.2, 33.4, -107.8, 35.3),
  zoom = 10,
  maptype = "stamen_terrain_background",
  color = "bw"
)

# Convert attribute location data to sf coordinates and change map projection
locations_sf <-
  st_as_sf(coord_df, coords = c("x", "y"), crs = 26912)
loc_trans <- st_transform(locations_sf, crs = 4326)
coord_4326_df <- data.frame(name = coord_df$name, st_coordinates(loc_trans))

# Extract edge list from network object
edgelist <- data.frame(get.edgelist(cibola_i)) %>% setNames(c("v1","v2"))

# add geometric info to the edge endpoints
edgelist <- left_join(edgelist,coord_4326_df, by = join_by(v1==name)) %>%
  setNames(c("v1","v2","X1","Y1"))
edgelist <- left_join(edgelist,coord_4326_df, by = join_by(v2==name)) %>%
  setNames(c("v1","v2","X1","Y1","X2","Y2"))

# Plot original data on map
ggmap(base_cibola, darken = 0.35) +
  geom_segment(
    data = edgelist,
    aes(
      x = X1,
      y = Y1,
      xend = X2,
      yend = Y2
    ),
    col = "white",
    alpha = 0.8,
    size = 1
  ) +
  geom_point(
    data = coord_4326_df,
    aes(X, Y, col = V(cibola_i)$Region),
    alpha = 0.8,
    size = 5,
    show.legend = FALSE
  ) +
  theme_void()
```

An alternative is that of transforming the entire graph into an sf object by assigning it a geometry.
This can be done using sfnetwork
```{r using sfnetwork}
# assign geometry attribute to cibola i
cibola_i_sf <- cibola_i
V(cibola_i_sf)$geometry <- loc_trans$geometry
cibola_i_sf<- as_sfnetwork(cibola_i_sf)

# quick plot
plot(cibola_i_sf)
```


# Exercises:
```{r space and betweenness}
# Create a data frame for the custom layout
coord_df <- data.frame(
  name = V(cibola_i)$name,
  x = V(cibola_i)$x, 
  y = V(cibola_i)$y)

# Define layout manually
ggraph(cibola_i, 
       layout = "manual", 
       x = coord_df$x,
       y = coord_df$y) +
  geom_edge_link(
    color ="gray"
  ) +
  geom_node_point(
    aes(size = igraph::betweenness(cibola_i),
        color = Region)
  ) +
  geom_node_text(
    aes(label = name)
  ) + 
  # select theme
  theme_graph()
```

With background map
```{r map and betweenness}
# Cibola map, spatial layout, node network as betwenness centrality 
# Plot original data on map
ggmap(base_cibola, darken = 0.35) +
  geom_segment(
    data = edgelist,
    aes(
      x = X1,
      y = Y1,
      xend = X2,
      yend = Y2
    ),
    col = "white",
    alpha = 0.8,
    size = 1
  ) +
  geom_point(
    data = coord_4326_df,
    aes(X, Y, col = V(cibola_i)$Region,  size = igraph::betweenness(cibola_i)),
    alpha = 0.8,
    show.legend = FALSE
  ) +
  theme_void()
```

```{r shortest path}
# find shortest path
Atsinna_UG494_path <- shortest_paths(cibola_i,
                                      from="Atsinna",
                                      to="UG494",
                                      output ="vpath")$vpath[[1]]

# transform into a df
Atsinna_UG494_path_df <- data.frame(v1 = Atsinna_UG494_path$name[1:3],
                                    v2 = Atsinna_UG494_path$name[2:4]) 

# add geometric info to the edge endpoints
Atsinna_UG494_path_df <- left_join(Atsinna_UG494_path_df,coord_4326_df, by = join_by(v1==name)) %>%
  setNames(c("v1","v2","X1","Y1"))
Atsinna_UG494_path_df <- left_join(Atsinna_UG494_path_df,coord_4326_df, by = join_by(v2==name)) %>%
  setNames(c("v1","v2","X1","Y1","X2","Y2"))

# Cibola map, spatial layout, node network as betwenness centrality 
# Plot original data on map
ggmap(base_cibola, darken = 0.35) +
  # add edges
  geom_segment(
    data = edgelist,
    aes(
      x = X1,
      y = Y1,
      xend = X2,
      yend = Y2
    ),
    col = "white",
    alpha = 0.8,
    size = 1
  ) +
  # add red segments to highlight the edges along the shortest path
  geom_segment(
    data = Atsinna_UG494_path_df,
    aes(
      x = X1,
      y = Y1,
      xend = X2,
      yend = Y2
    ),
    col = "red",
    alpha = 1,
    size = 1
  ) +
  # add nodes with colours depending on the region and size depending on betweenness centrality
  geom_point(
    data = coord_4326_df,
    aes(X, Y, col = V(cibola_i)$Region,  size = igraph::betweenness(cibola_i)),
    alpha = 0.8,
    show.legend = FALSE
  ) +
  # add labels of locations along the path
  geom_label(
    data = coord_4326_df[coord_4326_df$name %in%  Atsinna_UG494_path$name,],
    aes(X, Y, label = name),
    alpha = 1
  ) +
  # use a minimalistic theme
  theme_void() +
  # add title 
  labs(title = "Shortest path between Atsinna and UG494")
```

City network with map 
```{r city network}
# download stadiamap
base_city_map <- get_stadiamap(
  bbox = c(1.9, 41.3, 2.3, 41.6),
  zoom = 10,
  maptype = "stamen_terrain_background",
  color = "bw"
)

# read the same
city_network_p2<- read_graph("data/city_network.graphml", format ="graphml") # note: it adds an id attribute

# extract info on city locations
city_coord_df <- as.data.frame(vertex_attr(city_network_p2)[c("name","lon","lat")])

# Extract edge list from network object
city_edgelist <- data.frame(get.edgelist(city_network_p2)) %>% setNames(c("v1","v2"))

# add geometric info to the edge endpoints
city_edgelist <- left_join(city_edgelist,city_coord_df, by = join_by(v1==name)) %>%
  setNames(c("v1","v2","X1","Y1"))
city_edgelist <- left_join(city_edgelist,city_coord_df, by = join_by(v2==name)) %>%
  setNames(c("v1","v2","X1","Y1","X2","Y2"))

# Plot original data on map
ggmap(base_city_map, darken = 0.35) +
  geom_segment(
    data = city_edgelist,
    aes(
      x = X1,
      y = Y1,
      xend = X2,
      yend = Y2
    ),
    col = "white",
    alpha = 0.8,
    size = 1
  ) +
  geom_point(
    data = city_coord_df,
    aes(lon, lat, col = V(city_network_p2)$category, size = V(city_network_p2)$pop),
    alpha = 0.8,
    show.legend = FALSE
  ) +
  geom_text(
    data = city_coord_df,
    aes(label = name),
  ) +
  theme_void()
```
